Hunting experience shapes individual foraging specialization and predator-prey interactions in an online videogame

### 2023-05-03 manuscript/Chapter2-manuscript.Rmd
update MS fig1

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))` and the last one on `r toString(format(max(dat$date)))`. We cleaned and filtered the raw data to produce a dataset appropriate for our analyses. We analyzed only matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as it is supposed to be played. We then sampled players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)`-min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)`-min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific powers-gameplay mechanics), the game environment, the predator-player's experience along with its speed, and the average speed of the group of prey it encountered.

We recognize that we could have introduced a bias in our analyses since we sampled individuals who stayed in the game for at least 300 matches. For instance, these players may already be used to play videogames, and thus, could differ in the way they play at early stages of their experience in the game. To ensure that our sample was not biased, the videogame company provided a random sample of players that played between 20 and 50 matches, between 51 and 100 matches, and between 101 and 300 matches from the same timeframe as our current sample. We then took the first 20 matches played by these players, including those from our sampled population, and compared their average behaviour using a Bayesian hierarchical linear model. We found that neither of the four groups differed in their average speed (Table SX), which gives us confidence that our player population was not biased.

@@

We tested whether predators varied in the development of their expertise using three Bayesian generalized additive mixed models (GAMM) with thin plate regression splines. These models estimate the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function for all observations and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. Thus, this model assumes that predators do not share a common relationship between success and experience. We included match duration as a covariate in all three models.

Because a maximum of four prey can be captured in the game, and to control for overdispersion, we computed the three models using a modified version of the beta-binomial distribution implemented in "brms". Thus, hunting success was estimated as the probability of capturing the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link for all models.

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian priors for the game duration ($N(0,\ 1)$), the intercept ($N(0,\ 0.5)$), and the precision paremeter ($N(2,\ 0.5)$). We employed the default Student t priors for the smoothing parameter ($t(3,\ 0,\ 2.5)$). After fitting the three models, we proceeded to select the one with the best predictive accuracy using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model) and, by using a heterogeneous structure of the residuals, the intra-individual standard deviation of speed for every individual (which we call the dispersion part of the model). Thus, in the mean part of the model, we could estimate among individual differences in the foraging tactic used by predators (i.e. along the slow-fast continuum), while in the dispersion part of the model, we could estimate among individual differences in foraging specialization/flexibility (i.e. intra-individual variability). We followed the same structure for the preys' speed to estimate among individual differences in the mean and standard deviation of speed of the prey groups encountered by predators. For the mean part of the model, it allows us to test whether individual predators differed in the average speed of the groups of prey encountered across the study. For the dispersion part, it allows us to test whether individual predators differed in the (dis)similarity of the groups of prey encountered across the study (i.e. if resource heterogeneity was higher for some individuals compared to others). For hunting success, we only modeled the mean part of the equation to estimate among individual differences in mean prey capture. For the predator and the prey's speed, we controlled for the average rank of the prey that they encountered as proxy for prey skills on both the mean and dispersion part of the equation. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). We used the mean rank of the four prey and applied a square-root transformation. For the predator's hunting success, we controlled for the square-root of the duration of the match.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, the correlations enables us to estimate, for example, whether predators that encountered prey that were faster/slower adopted a slower/faster (or more specialized/flexible) tactic, or if slower/faster predators had lower/greater hunting success. As a complement, the model also estimated correlations among the different levels of experience (e.g. $cor(\sigma_{id_{m}\ pred\ novice}\ \sigma_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each of three developmental stages. We present the character-state analyses as additional material in the Appendix S2 as they were not part of our initial objectives.

@@

We parameterized equations 1.1 to 2.2 with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation 3 (predator hunting success). We used Gaussian priors for the prey rank ($N(2.5,\ 1)$) and game duration ($N(10,\ 2)$) on the mean part of the model (equations 1.1, 2.1, and 3), and the default priors in "brms" for the dispersion part (equations 1.2 and 2.2). We used the default Student t priors in "brms" for the intercepts on the mean and dispersion part of the model. For hunting success, we applied a Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a half-Gaussian prior ($N(0,\ 1)$) on the random effects for the mean part of the model (equations 1.1, 2.1, and 3), and the default "brms" Student t prior for the dispersion part of the model (equations 1.2 and 2.2).

@@

Our results confirm that individual predators display substantial differences in the development of their expertise, as the best model included group-level smoothers exclusively (Table I). For instance, some predators steadily increased (Figure 1A, Appendix S1: Figure S1A), while others decreased (Figure 1A, Appendix S1: Figure S1B), their hunting success with greater experience, and others whose success did not change with experience (Figure 1A). Another pattern that emerged was that some individuals optimized their success at an experience level that was below what we considered "advanced" (i.e. >300 matches played).

We predicted that the relationship between hunting success and experience would vary among predators because they should differ in the type of prey that they encounter. However, after controlling for the mean speed of the prey group within a match, we did not find evidence confirming this prediction, because differences among predators remained stable (Figure 1B).

parler du % de joueurs qui ont eu une augmentation vs % qui ont eu une diminution

@@

**Figure 2.** Posterior median differences and credible intervals comparing the parameter values predicted by the MDHGLM between each predator experience level. Each panel represent the experience levels that are compared. The parameter being compared is displayed on the y axis (e.g. whether the standard deviation of predator ID on the mean differs when the population was intermediate vs novice), and the difference is displayed on the x axis.

>>>>

The videogame company provided data that spanned a period of 6 months of gameplay recorded for every player. The first recorded match was played on `r toString(format(min(dat$date)))` and the last one on `r toString(format(max(dat$date)))`. We analyzed only matches where players did not know each other (i.e. "Online" mode). We filtered any matches where players were inactive, such as when speed values were equal to, or very close to, zero. Moreover, we used our knowledge of the game to remove any matches where players were potentially hacking, or not playing the game as it is supposed to be played. We then sampled players that played 300 matches or more, and monitored all their matches from the first to a maximum of 500 matches.

Our population consists of `r toString(format(length(unique(dat$predator_id))))` predator players with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches. The predator-players' experience varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)`-min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)`-min). The following information is collected and reported for every match : the player's anonymous ID, its avatar (i.e. the predator character chosen with its specific skill-gameplay mechanics), the game environment, the predator-player's experience along with its speed, and the average speed of the group of prey it encountered.

We recognize that we could have introduced a bias in our analyses since we sampled individuals who stayed in the game for at least 300 matches. For instance, these players may already be used to play videogames, and thus, could differ in the way they play at early stages of their experience in the game. To ensure that our sample was not biased, we compared a random sample of players that played between 20 and 50 matches, between 51 and 100 matches, and between 101 and 300 matches from the same timeframe with our sampled population. We then took the first 20 matches played by these players, including those from our sampled population, and compared their average behaviour using a Bayesian hierarchical linear model. We found that neither of the four groups differed in their average speed (Appendix 2: Table S1-S2), which gives us confidence that our player population was not biased.

@@

We tested whether predators varied in the development of their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin plate regression splines. These models estimate the relationship between hunting success (i.e. number of prey captured) and the predators' cumulative experience (i.e. number of matches played before the current match). We parametrized the models following the method of @Pedersen.etal2019. The first model was the simplest, with a common global smoothing function for all observations and random intercepts for the predator ID. In this model, we assume that individuals have the same development of expertise, with the model estimating a trend for the average individual (i.e. global smoother). For the second model, we included varying group-level smoothers for the predator ID. In this model, we assume that individuals share a similar relationship between success and experience, but that this relationship can vary (e.g. individual 1 has a steeper curve than individual 2). This enabled us to test whether predators differed in the development of their expertise. In the third model, we kept the group-level smoothers for the predators, but removed the global smoother. This allows each individual to have a unique relationship between success and experience without penalization by the global smoother. Thus, this model assumes that predators do not share a common relationship between success and experience. The fourth and fifth model were a reproduction of the second and third model respectively, where we also included the prey group's average speed to assess their effect on the relationship between success and experience. We included match duration as a covariate in all five models.

Because a maximum of four prey can be captured in the game, and to control for overdispersion, we computed the five models using a modified version of the beta-binomial distribution implemented in "brms". Thus, hunting success was estimated as the probability of capturing the four prey ($\mu_i$), drawn from a Beta distribution ($Beta(\mu_i, \phi)$) with mean ($\mu \in [0,1]$) and precision ($\phi > 0$) parameters. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor, while the precision parameter ($\phi$) was estimated with an identity link for all models.

We used the default number of basis functions (K) in "brms" for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id})$). We used Gaussian weakly informative priors for the intercept ($N(0,\ 0.5)$) and the global trend of cumulative experience ($N(0,\ 2)$). Based on previous results [@FraserFranco.etal2022], we defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5)$), a positive Gaussian prior ($N(1,\ 0.5)$) on the game duration because longer trials lead to greater success, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5)$) because encountering faster prey is associated with lower success in this system. We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0, 0.5)$). We compared the models' predictive accuracies using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@Vehtari.etal2017;@Piironen.Vehtari2017;@Vehtari.etal2022].

@@

For each level of experience (i.e. novice, intermediate, advanced), the model estimated the mean speed of every individual (which we call the mean part of the model), and by explicitly modelling the residuals, the intra-individual standard deviation of speed for every individual (which we call the dispersion part of the model). In the mean part of the model, we could estimate among individual differences in the foraging tactic used by predators (i.e. along the slow-fast continuum), while in the dispersion part of the model, we could estimate among individual differences in foraging specialization/flexibility (i.e. intra-individual variability, IIV). We followed the same structure for the preys' speed to estimate among individual differences in the mean and standard deviation of speed of the groups of prey encountered by predators. For the mean part of the model, it allows us to test whether individual predators differed in the average speed of the groups of prey encountered across the study. For the dispersion part, it allows us to test whether individual predators differed in the (dis)similarity of the groups of prey encountered across the study (i.e. if resource heterogeneity was higher for some individuals compared to others). For hunting success, we modeled the mean part of the equation to estimate among individual differences in mean prey capture. For the predator and the prey's speed, we controlled for the average rank of the prey that they encountered as proxy for prey skills on both the mean and dispersion part of the equation. The ranking system in *DBD* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). We used the mean rank of the four prey and applied a square-root transformation. For the predator's hunting success, we controlled for the square-root of the game duration.

We estimated the correlation between the individual random effect on the mean and on the dispersion within each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ pred})$) and among each trait (e.g. $cor(\sigma_{id_{m}\ pred}\ \sigma_{id_{d}\ prey})$) within a level of experience. In the first case, this allows us to assess, for example, whether predators that are on average slower/faster are more specialized or flexible. In the second case, it enables us to estimate, for example, whether predators that adopted a slower/faster (or more specialized/flexible) tactic encountered prey that were faster/slower, or if slower/faster predators had lower/greater hunting success. As a complement, the model also estimated correlations among the different levels of experience (e.g. $cor(\sigma_{id_{m}\ pred\ novice}\ \sigma_{id_{d}\ pred\ intermediate})$). The latter "character-state" approach [@Via.Lande1985;@viaAdaptivePhenotypicPlasticity1995], enables us to investigate whether predators express similar behaviours across each of three developmental stages. We present the character-state analyses in the Appendix S4 as they were not part of our initial objectives.

@@

We parameterized equations 1.1 to 2.2 with a Gaussian distribution family (predator and prey speed) and used a beta-binomial distribution family for equation 3 (predator hunting success). We used weakly informative Gaussian priors for the prey rank ($N(0,\ 1)$) and game duration ($N(0.5,\ 0.5)$) on the mean (equations 1.1, 2.1, and 3) and on the dispersion part of the model (equations 1.2 and 2.2). We used a weakly informative Gaussian prior ($N(0,\ 1)$) for the intercepts on the mean and dispersion part of the model. For hunting success, we applied a positive Gaussian prior on the precision parameter ($N(2,\ 1)$). For every predicted variable, we applied a weakly informative half-Gaussian prior ($N(0,\ 1)$) on the random effects for the mean part and the dispersion part of the model. We used a regularizing $LKJ(2)$ prior on the correlation matrix.

After inspection of the results, we found that many parameters values were in the direction of our predictions, but they were largely driven by the very slow speed of one individual predator. Removing this individual reduced the estimates by a significant margin. To remain conservative, we thus present the results of the model without this individual in the main text, and provide the results of the model including it in the Appendix S3.

@@

Out of all five models, the two that best predicted the data, with an almost identical expected log pointwise density (elpd), were the model with group-level smoothers exclusively while controlling for prey speed and the model including a global smoother of predator experience along with group-level smoothers and prey speed (Table 1). Thus, prey speed was an important predictor of the predators' hunting success.

Our results confirm that predators differ substantially in the relationship between hunting success and cumulative experience (i.e. expertise: Figure 1A-B). For instance, the success of some predators steadily increased with experience (Figure 1B, Appendix S1: Figure S1A) while others showed a decrease in success with experience (Figure 1B, Appendix S1: Figure S1B). Overall, `r toString(scales::percent(191/253))` of the population displayed a >10% increase in hunting success, `r toString(scales::percent(37/253))` displayed a >10% decrease in hunting success, and the remaining `r toString(scales::percent(25/253))` had a lesser change in success ranging between -10% and +10%.

We predicted that the relationship between hunting success and experience would vary among predators because they should differ in the type of prey that they encounter. After controlling for the mean speed of the prey group within a match, we did not find evidence confirming this prediction, because differences among predators remained important (Figure 1A-B). However, many individual curves appear to be shifted upwards in the model where we control for prey speed (Figure 1B) vs in the model where we do not control for prey speed (Figure 1A). Moreover, in the models where we did not account for prey speed, the global trend of predator experience was nearly flat (results not shown). Yet, we see that there is an increase up to 13% in the probability of capturing all prey with experience in the population when we account for prey speed (Figure 1C). The maximum is attained around match 340 (i.e. advanced state). Prey speed was thus important in mediating the population and individual-level relationship between success and experience, but does appear to influence among individual differences in this relationship.

@@

**Figure 2.** Posterior median differences and credible intervals comparing the parameter values predicted by the MDHGLM between each predator experience level. Each panel represent the experience levels that are compared. The parameter being compared is displayed on the y axis (e.g. whether the standard deviation of predator ID on the mean differs when the population was intermediate vs novice), and the difference is displayed on the x axis. *développer pour expliquer mean et sigma réfère à quoi. prendre la légende du tableau en appendix 1*
